---
title: "Predict Correct Use of Barbells"
author: "Frank Jung"
date: "14 October 2015"
output:
  html_document:
    highlight: monochrome
    toc: yes
---

```{r initialisation, echo=FALSE, message=FALSE, warning=FALSE}
require(knitr, quietly = TRUE)
require(caret, quietly = TRUE)
require(dplyr, quietly = TRUE)
require(ggplot2, quietly = TRUE)
require(rfUtilities, quietly = TRUE)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r load, echo=FALSE}
# load raw data
raw <- read.csv("data/pml-training.csv", header = TRUE, na.strings = c("NA", "#DIV/0!"), stringsAsFactors = FALSE)

# set classe as a factor
raw$classe <- factor(raw$classe)
```

## Summary

This report describes the algorithm used to predict how people did an exercise 
of lifting a dumbbell. This is a classification problem where predictors are
used to determine one of the five [outcomes](#goal). This report will:

* describe how the model was built
* what choices were made for this model
* how the model was cross validated
* what the expected out of sample error is
    
Links to the data and further information of the research can be the
[Reference](#references)

## Background

Using devices such as [Jawbone Up](https://jawbone.com/up), [Nike
FuelBand](http://www.nike.com/), and [Fitbit](https://www.fitbit.com) it is now
possible to collect a large amount of data about personal activity relatively 
inexpensively. These type of devices are part of the quantified self movement – 
a group of enthusiasts who take measurements about themselves regularly to 
improve their health, to find patterns in their behavior, or because they are 
tech geeks. One thing that people regularly do is quantify how much of a 
particular activity they do, but they rarely quantify how well they do it. 

## Goal

Our goal is to use data from accelerometers on the belt, forearm, arm, and 
dumbbell of 6 participants to predict the manner in which people did an 
exercise. They were asked to perform dumbbell lifts correctly 
and incorrectly in 5 different ways:

```{r classe-table}
classe.id <- c(levels(raw$classe))
classe.desc <- c("exactly according to the specification", 
                 "throwing the elbows to the front",
                 "lifting the dumbbell only halfway",
                 "lowering the dumbbell only halfway",
                 "and throwing the hips to the front")
classe.table <- data.frame(Classe = classe.id, Description = classe.desc)
kable(classe.table)
```

The goal of this project is to predict the manner in which people did an 
exercise. This is recorded in the "classe" variable of the training set.



## Exploratory Analysis


Actions:

(1) Will to convert Excel's '#DIV/0!' to NA.
(2) Do not include row number field ``X``as not required regression model.
(3) Factors

```{r partition}
# split into train (70%) and test (30%) on outcome (classe)
rawindex <- createDataPartition(raw$classe, p = 0.7, list = FALSE)
training <- raw[rawindex,]
testing <- raw[-rawindex,]
```

```r
# ignore columns that are more than 95% empty (i.e. NA):
nas.perc <- as.integer(0.95 * nrow(raw))
nas <- sort(apply(raw, 2, function(x) length(which(is.na(x)))), decreasing = TRUE)
bad.names <- sort(names(nas[nas >= 19216])); bad.names
good.names <- setdiff(names(training), bad.names); good.names

# exclude columns that do not aid in prediction (or are an outcome)
train.names <- grep(paste("classe", "window", "user_name", "X", "_timestamp", sep = "|"), good.names, value = TRUE, invert = TRUE); train.names

# use these column names to generate training formula
train.formula <- as.formula(paste("classe ~ ", paste(train.names, collapse = "+"))); train.formula
```

Outcome = classe


Split on time series or num_windows field:
See http://topepo.github.io/caret/splitting.html

```r
# select just num_window and classe from data frame
train.unique <- unique(subset(training, select = c("num_window", "classe")))

# how many unique combinations of num_window and classe?
nrow(train.unique)
# [1] 858

# how many unique windows?
length(unique(training$num_window))
# [1] 858

# so can use num_windows to create data partitions
```

# Appendices

## References

* Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. [Qualitative Activity Recognition of Weight Lifting Exercises](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201). Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. [PDF](http://groupware.les.inf.puc-rio.br/public/papers/2013.Velloso.QAR-WLE.pdf)

## Session Information

```{r environment}
sessionInfo()
```

## GitHub

* [Source](https://github.com/frankhjung/predmachlearn-033)
* [HTML Documentation](https://frankhjung.github.io/predmachlearn-033)
