---
title: "Predict Correct Use of Dumbbells"
author: "Frank Jung"
date: "17 October 2015"
output:
  html_document:
    highlight: monochrome
    toc: yes
---

```{r initialisation, echo=FALSE, message=FALSE, warning=FALSE}
require(knitr, quietly = TRUE)
require(caret, quietly = TRUE)
require(dplyr, quietly = TRUE)
require(ggplot2, quietly = TRUE)
require(rfUtilities, quietly = TRUE)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r load, echo=FALSE}
# load raw data
raw <- read.csv("data/pml-training.csv", header = TRUE, na.strings = c("NA", "#DIV/0!"), stringsAsFactors = FALSE)

# set classe as a factor
raw$classe <- factor(raw$classe)
```

## Summary

This report describes the algorithm used to predict how people did an exercise
of lifting a dumbbell. This is a classification problem where predictors are
used to determine one of the five [outcomes](#goal). This report will:

* describe how the model was built
* what choices were made for this model
* how the model was cross validated
* what the expected out of sample error is

Links to the data and further information of the research can be the
[references](#references)

## Background

Using devices such as [Jawbone Up](https://jawbone.com/up), [Nike
FuelBand](http://www.nike.com/), and [Fitbit](https://www.fitbit.com) it is now
possible to collect a large amount of data about personal activity relatively
inexpensively. These type of devices are part of the quantified self movement –
a group of enthusiasts who take measurements about themselves regularly to
improve their health, to find patterns in their behavior, or because they are
tech geeks. One thing that people regularly do is quantify how much of a
particular activity they do, but they rarely quantify how well they do it.

## Goal

We wish to predict the the manner in which people did an exercise from data
collected from accelerometers on the belt, forearm, arm, and dumbbell of 6
participants. They were asked to perform dumbbell lifts correctly and
incorrectly in 5 different ways:

```{r classe-table}
classe.id <- c(levels(raw$classe))
classe.desc <- c("exactly according to the specification",
                 "throwing the elbows to the front",
                 "lifting the dumbbell only halfway",
                 "lowering the dumbbell only halfway",
                 "and throwing the hips to the front")
classe.table <- data.frame(Classe = classe.id, Description = classe.desc)
kable(classe.table, caption = "Classes of exercise activity")
```

## Exploratory Analysis

The ``classe`` field is a categorical outcome, which we will treat as a factor.

The data contains a large number (`r ncol(raw)`) of columns. However, many of
these columns contain little or no information or contain an Excel ``#DIV/0!``
conversion error. Consequently, we will exclude from the prediction formula
all columns with a 95% or higher missing data.

A number of columns will also be excluded since they do not provide value to
prediction. They are:

| Column | Reason |
|--------|--------|
| ``X`` | Row index |
| ``*_window``, ``*_timestamp*`` | measurements were made in time segments with overlap so they can be treated as discrete and not time dependent |
| ``user_name`` | who performed the action is not relevant as all were supervised to ensure consistent actions |

Table: Additional columns to ignore

## Choice of Prediction Algorithm

A good performing prediction algorithm for classification problems is [Random
Forest](https://en.wikipedia.org/wiki/Random_forest). It is resilient to
outliers but is affected by multi-collinearity, which will be tested for, later.

The predictors used in the formula, were tested using
[rfUtilities](http://cran.r-project.org/package=rfUtilities). No multi-collinear
variables were identified.

### Data Partitioning

The [training data](#data) was partitioned into two partitions:

* training (70%) - used to train model
* testing (30%) - used for cross-validation, and error estimation

```{r partition, echo = TRUE}
# split into train (70%) and test (30%) on outcome (classe)
rawindex <- createDataPartition(raw$classe, p = 0.7, list = FALSE)
training <- raw[rawindex,]
testing <- raw[-rawindex,]
```

### Training Formula

The model formula is composed of all fields, **except**

* columns that are more than 95% empty (i.e. NA)
* non-predictive columns identified in [Table: Additional columns to ignore](#exploratory-analysis)

Applying these conditions leads to the formula with an outcome ``classe`` of:

```{r formula, fig.caption="Formula"}
# ignore columns that are more than 95% empty (i.e. NA):
nas.perc <- as.integer(0.95 * nrow(raw))
nas <- sort(apply(raw, 2, function(x) length(which(is.na(x)))), decreasing = TRUE)
bad.names <- sort(names(nas[nas >= 19216]))
good.names <- setdiff(names(training), bad.names)

# exclude columns that do not aid in prediction (or are an outcome)
train.names <- grep(paste("classe", "window", "user_name", "X", "_timestamp", sep = "|"), good.names, value = TRUE, invert = TRUE)

# use these column names to generate training formula
train.formula <- as.formula(paste("classe ~ ", paste(train.names, collapse = "+"))); train.formula
```

This will appear below in training as the variable ``train.formula``.

### Training Model

The Random Forest was trained using [caret](http://caret.r-forge.r-project.org/
) package with:

```r
model <- train(train.formula, data = training, method = "rf")
```

```{r train}
# model using random forest
if (file.exists("data/model-rf.rds")) {
    model <- readRDS("data/model-rf.rds")
}
```

### Check Multi-Collinearity

Multi-collinearity of the model was tested with:

```{r, echo=TRUE}
multi.collinear(dplyr::select(training, one_of(train.names)))
```

### Variable Importance

Below is an abbreviated list of variable importance:

```{r importance}
varImp(model)

# full list with
# importance(model$finalModel)

# plot with
# par(mfrow = c(1, 1), family = "sans", cex = 0.7)
# varImpPlot(model$finalModel, n.var = 20, main = "Variable Importance")
```

## Testing

The model was cross-validated against the test data with:

```{r predict, echo = TRUE}
test.predict <- predict(model, newdata = testing)
```

### Accuracy and Cohen Kappa

The accuracy and Cohen [Kappa](https://en.wikipedia.org/wiki/Cohen%27s_kappa)
statistics are:

```{r accuracy}
cm <- confusionMatrix(data = test.predict, reference = testing$classe)
# ignore AccuracyNull(5) and McnemarPValue(7)
overall.stats <- data.frame(Value = round(cm$overall[-c(5,7)], 3))
kable(x = overall.stats, caption = "Overall Statistics")
```

### Confusion Matrix

The predictions [confusion
matrix](https://en.wikipedia.org/wiki/Confusion_matrix) is:

```{r confusionMatrix}
byclass <- round(cm$byClass, 3)
kable(byclass, caption = "Confusion Matrix")
```

## Validation

The final model was run against the validation data yielding the following
predictions:

```r
validation <- read.csv(
    "data/pml-testing.csv", header = TRUE,
    na.strings = c("NA", "#DIV/0!"), stringsAsFactors = FALSE
)
predictions <- predict(model, newdata = validation)
saveRDS(predictions, "data/predictions-rf.rds")
```

The predictions are then processed using the [script](scripts) ``pred.R``

## Appendices

### References

* Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. [Qualitative Activity Recognition of Weight Lifting Exercises](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201). Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. [PDF](http://groupware.les.inf.puc-rio.br/public/papers/2013.Velloso.QAR-WLE.pdf)

### Data

Data for this project was sourced from:

* Training data, https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
* Validation data, https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

### Scripts

| Name | Description |
|------|-------------|
| getdata.R | download raw CSV data from remote sources |
| model-rf.R | train random forest model |
| pred.R | prepare predictions for submission |

### GitHub

* [Source](https://github.com/frankhjung/predmachlearn-033)
* [HTML Report](https://frankhjung.github.io/predmachlearn-033/project.html)
